Improving Adaptive Seamless Designs through Bayesian optimization

### 2020-12-21 main.tex
Updates from Overleaf

\title[Bayesian optimization of Adaptive Seamless Designs]{Optimizing the power of adaptive seamless designs using Bayesian optimization} %(please use sentence case, e.g. Score tests for exploring complex models)
%% Information for the first author.
\author[Jakob Richter {\it{et al.}}]{Jakob Richter\footnote{Corresponding author: {\sf{e-mail: richter@statistik.tu-dortmund.de}}}\inst{,1}} 
\address[\inst{1}]{Fakultät Statistik, Technische Universität Dortmund, 44221 Dortmund}
%%%%    Information for the second author
\author[dd]{Tim Friede\inst{2}}
\address[\inst{2}]{Institut für Medizinische Statistik, Universitätsmedizin Göttingen, 37073 Göttingen}
%%%%    Information for the third author
\author[]{Jörg Rahnenführer\inst{1}} %please provide full author names. Middle names should be indicated by initials only, i.e. Henry J. James
%%%%    \dedicatory{This is a dedicatory.}
\Receiveddate{zzz} \Reviseddate{zzz} \Accepteddate{zzz}

@@

% max 250 words
\begin{abstract}
In clinical trials, planning patient group sizes and test methods is a crucial task.
A common goal is to maximize the test power, given a set of treatment and effect sizes under the null hypothesis.
From a wide range of possible alternatives we want to select the best plan in a short period of time, to enable quick decisions.
The standard approach to simulate the power for each single setup can become too time-consuming, when the number of possible setups becomes very large.
Then, either large computational resources are required or an exhaustive exploration of all possible setups takes a very long time.
In this work we propose to use Bayesian optimization to quickly find a clinical trial setup with high power from a large number of setups options.
We demonstrate the effectiveness of our approach by optimizing the power of adaptive seamless designs for different sets of treatment effect sizes.
We compare the Bayesian optimization approach with an exhaustive evaluation of all candidate designs.
The optimization approach finds competitive setups in a fraction of the time.
\end{abstract}

@@

% FIXME: Do we really take in this Assurance?
Assurance: \cite{stallard_optimal_2009} in diskussion abhandeln

@@

We start by providing some more details on the study design considered (Sections \ref{ssec:two_stage_design_with_treatment_selection} and \ref{ssec:selection_rules}) and then introduce briefly a framework for optimizing such designs, namely clinical scenario evaluation (CSE) in Section \ref{ssec:clinical_scenario_evaluation_for_asd}. In Section \ref{ssec:simulating_asd} we make some brief comments on the simulation model.

\subsection{Two-stage design with treatment selection}
\label{ssec:two_stage_design_with_treatment_selection}
Here we consider a particular type of adaptive seamless designs, namely two-stage designs with treatment (or more specifically dose) selection. The trial starts in stage 1 with $K$ experimental treatments or doses of one experimental treatment which are compared to a common control. Interest is in assessing the treatment effects $\beta_k$ with $k=1,\dots, K$ comparing experimental treatment $k$ vs. control and testing the null hypotheses $H_{0,k}: \beta_k=0$ against the alternative  $H_{1,k}: \beta_k>0$ (larger $\beta_k$ being better). For the purpose of treatments selection, test statistics $Z_{1,k}$ based on data from the first design stage are considered with larger $Z_{1,k}$ indicating more beneficial treatments. These can use data on the primary outcome used for testing or any other outcome that is available at the time of the interim analysis. In practical applications, early outcomes are often used to inform interim decisions, since the primary endpoint might take a longer time to observe. In the interim analysis, treatments are selected for continuation into the second stage; the set of selected treatments is denoted by $\mathcal{S}$. The selection rules used here are described in Section \ref{ssec:selection_rules}. In the final analysis, the closed test principle is applied to account for the multiple hypotheses. The intersection hypotheses are tested by combining stagewise $p$-values for the respective intersection hypotheses using a prespecified combination function. Here we use the inverse normal combination function~\citep{lehmacher_adaptive_1999}:
\begin{equation}
  \frac{1}{\sqrt{K}} \sum_{k=1}^K \Phi^{-1}(1-p_k) \,
\end{equation}
where $\Phi^{-1}$ denotes the quantile function of the standard normal distribution.

\subsection{Selection rules}
\label{ssec:selection_rules}
In practice, the totality of the data would be suggested by a data monitoring committee (DMC) to make a recommendation regarding the treatment selection. However, it is advisable for the sponsor to define a selection rule, although in comparison fairly simplistic, to provide some guidance for the DMC regarding sponsor's preferences. Furthermore, formal selection rules are indispensable to conduct simulation studies, e.g. to explore sample size and power.

\paragraph{k-Rule} %best k, evtl. anderer Name?

\paragraph{Epsilon Rule}

\paragraph{Threshold Rule}

\subsection{Simulating ASD}
\label{ssec:simulating_asd}
The R-package \texttt{asd} by~\citet{parsons_software_2011} simulates the test statistics and not individual participant data.

@@

\begin{figure}[ht]
\centering
\begin{tikzpicture}[auto,node distance=1cm,align=center,text width = 3cm, inner sep = 0.1cm, font=\sffamily \scriptsize, outer sep= 0cm]
  \node[draw, ellipse, minimum height=1cm, text width=2.5cm] (ClinicalScenarios) at (0,0) {
    \textbf{Clinical scenarios}\\
    Simulation Studies};
  \node[name=block, rectangle split, rectangle split parts=2, draw, above = of ClinicalScenarios] (UnknownPoints)
  { \begin{itemize}[leftmargin=.3cm, topsep=0cm,itemsep=0cm,partopsep=0cm, parsep=0cm]
      \item Treatment effect on early interim outcome
      \item Treatment effect on late disability outcome
    \end{itemize}
    \nodepart{second} Unknown
  };
  \node[name=block, rectangle split, rectangle split parts=2, draw, anchor = north east] (HistoricalDataPoints) at (UnknownPoints.north west)
  { \begin{itemize}[leftmargin=.3cm, topsep=0cm,itemsep=0cm,partopsep=0cm, parsep=0cm]
      \item Correlation between early interim and late disability outcomes
    \end{itemize}
    \nodepart{second} Historical data
  };
  %\node[draw, inner xsep=1em, inner ysep=1em, above left = of ClinicalScenarios, fit = (HistoricalDataPoints) (UnknownPoints)] (boxDiseaseSpecific) {};
  \node[draw, text width = 6.2cm, anchor = south west] (DiseaseSpecific) at (HistoricalDataPoints.north west) {\textbf{Disease-specific features}};
  %\node[draw, above right = of ClinicalScenarios] (DesignOptions) {Design options};
  \node[name=block, rectangle split, rectangle split parts=2, draw, anchor = north west] (ConstrainedPoints) at ([xshift = 4cm]HistoricalDataPoints.north east)
  { \begin{itemize}[leftmargin=.3cm, topsep=0cm,itemsep=0cm,partopsep=0cm, parsep=0cm]
      \item Number of test treatments
      \item Total subjects in trial
      \item Recruitment rate
    \end{itemize}
    \nodepart{second} Constrained by health care environment/infrastructure
  };
  \node[name=block, rectangle split, rectangle split parts=2, draw, anchor = north west] (CombinationPoints) at (ConstrainedPoints.north east)
  { Adaptive seamless design
    \begin{itemize}[leftmargin=.3cm, topsep=0cm,itemsep=0cm,partopsep=0cm, parsep=0cm]
      \item time point of interim analysis
      \item treatment selection rules at interim Conventional phase 2/3
      \item split of resources between phases
      \item end oh phase futility criteria
    \end{itemize}
    \nodepart{second} Combination of options
  };
  \node[draw,text width = 6.2cm, anchor = south west] (DesignOptions) at (ConstrainedPoints.north west) {\textbf{Design Options}};
  \node[draw, below = 0.5cm of ClinicalScenarios] (DesignPerformance) {
    \textbf{Design performance} \\ e.g.\ statistical power};
  \node[draw, right = of DesignPerformance, text width = 4cm, node distance = 2cm] (Evaluation) {
    \textbf{Clinical Scenario Evaluation}\\
    Design performance measures evaluated across a wide range of clinical scenarios};
  \draw[->] (HistoricalDataPoints) -- (ClinicalScenarios);
  \draw[->] (UnknownPoints) -- (ClinicalScenarios);
  \draw[->] (CombinationPoints.223) -- (ClinicalScenarios);
  \draw[->] (ConstrainedPoints) -- (ClinicalScenarios);
  \draw[->] (ClinicalScenarios) -- (DesignPerformance);
  \draw[->] (DesignPerformance) -- (Evaluation);
\end{tikzpicture}
%Friede et al (2010) DIJ, Figure 2 https://link.springer.com/content/pdf/10.1177/009286151004400507.pdf? https://link.springer.com/content/pdf/10.1177/009286151004400607.pdf
\caption{Application of the refined clinical scenario evaluation framework to adaptive seamless designs in progressive multiple sclerosis.}
\label{fig:friede_refinement_2010_fig2}
\end{figure}

@@

\clearpage

@@

The strategies for treatment selection for the second stage in the two-phase trial design and their corresponding parameters that are used in the simulation study are given in Table~\ref{tab:selection_strategies} and Table~\ref{tab:search_space}, respectively.
The strategies follow the concept of adaptive seamless designs discussed in Section~\ref{sec:adaptive_seamless_designs}. 
The strategies 1, 2, 3, and \emph{all} select the $k$ best strategies according to their respective test statistics in the first stage (with \emph{all} representing all $k=4$ treatments besides the control). 
The strategies \emph{eps} and \emph{thresh} represent strategies with a flexible number of treatments taken to the second stage. 
In Table~\ref{tab:search_space} the ranges for these two parameters and for the proportion of patients (ratio $r$) allocated to the second stage are given.
\begin{table}[h]
  \caption{Overview of selection strategies to determine which trial arms are kept for second stage}
  \label{tab:selection_strategies}
  \centering
  \begin{tabular}{ll}
  \hline
  Notation       & Procedure of selection strategy  \\
  \hline
  \emph{all}     & Select all arms  \\
  $\{1,2,3\}$    & Select 1, 2 or 3 arms, respectively, according to maximal test statistic \\
  \emph{eps}     & Select all arms with test statistic less than $\epsilon$ (epsilon) below maximal test statistic \\
  \emph{thresh}  & Select all arms with test statistic larger than the value $\tau$ (threshold) \\
  \hline
  \end{tabular}
\end{table}
\begin{table}[h]
  \caption{Parameters for the selection strategies representing the search space of trial designs}
  \label{tab:search_space}
  \centering
  \begin{tabular}{ll}
  \hline
  Parameter                        & Range \\
  \hline
  Selection strategy               & $\{\text{all}, 1, 2, 3, \text{eps}, \text{thresh} \}$ \\
  $r$ (ratio)                      & $(0,1)$ \\
  $\epsilon$ (only for \emph{eps}) & $[0,4]$ \\
  $\tau$ (only for \emph{thresh})  & $[0,10]$ \\
  \hline
  \end{tabular}
\end{table}

@@

\paragraph{Grid}
We conduct an \emph{exhaustive grid search} with a resolution of 25 points per real-valued dimension and one point for each categorical dimension.
Referring to our application, for each selection strategy we evaluate a numerical grid with a resolution of 25.
All 6 selection strategies have the numeric parameter $r$.
The strategies \emph{epsilon} and \emph{thresh} both have an additional second parameter, $\epsilon$ or $\tau$, respectively. 
This results in a total of $4 \cdot 25 + 2 \cdot 25 \cdot 25 = 1350$ design configurations that are evaluated for one scenario.
We call this approach \textbf{Grid}.
Further, we also consider a subset of this grid of 7 points per dimensions, resulting in $4 \cdot 7 + 2 \cdot 7 \cdot 7 = 126$ design configurations, which we will call \textbf{Grid Small}.
Both grids are evaluated with 20 stochastic replications to estimate the variance of the estimated power.

@@

\paragraph{BO (Model-based optimization)}
For applying BO, we first define how the surrogate model that predicts the outcome of $f$ for unknown values of $\theta$ (parameter configurations) is generated.
As regression model for the surrogate we choose Kriging (also called Gaussian process regression) and use the implementation in the R-package \texttt{DiceKriging}~\citep{roustant_dicekriging_2012}, because it is known to work well for fairly low dimensional search spaces.
We configure Kriging to apply the Mattern$\frac{5}{2}$ kernel with an estimated \emph{nugget effect} to account for the noisy response of $f$ and without scaling the input variables to $[0,1]$.

@@

The whole BO procedure is repeated $20$ times to estimate the variance of the optimal found solution.

@@

The algorithms used in this study are implemented in R.
For the Bayesian optimization, the R-package \texttt{mlrMBO}~\citep{bischl_mlrmbo_2017} is used.
For the simulation of clinical trials, the function \texttt{treatsel.sim} of the R-package \texttt{asd}~\citep{parsons_software_2011} is used, with parameter values given in Table~\ref{tab:par_implement}.
\begin{table}[h]
  \caption{Parameter choices for the R function \texttt{treatsel.sim}}
  \label{tab:par_implement}
  \centering
  \begin{tabular}{lll}
  \hline
  Meaning of parameter & Parameter & Value \\
  \hline
  Number of simulation iterations & \texttt{nsim} & $1000$ \\
  Correlation between early and final outcomes & \texttt{corr} & $0.4$ \\
  Test level & \texttt{level} & $0.025$ \\
  Vector of treatment numbers for determining power & \texttt{ptest} & $(3,4)$ \\
  \hline
  \end{tabular}
\end{table}
The vector of treatment numbers for determining power \texttt{ptest} counts rejections of 3 or 4 hypotheses ???.
Note that the parameter \texttt{nsim} denotes the internal simulation iterations of $f$. 
The ten additional stochastic repetitions we conduct for the \emph{grid search} and \emph{BO} are additional replicates.

@@

\subsection{Main results}

In this chapter we present the results of the four optimization strategies: exhaustive \emph{Grid} search, grid search with a reduced resolution (\emph{Grid Small}), \emph{BO} and \emph{BO Grid},
We focus our comparison on independent, replicated evaluations of the configurations proposed by each optimizer.
The replicated evaluations give us an unbiased estimate of the performance ($y_{\text{valid}}$) of each optimizer.

For this study we consider 12 scenarios.
The scenarios are generated by combining the four different effect sets (see Table~\ref{tab:table_effect_names}) with the three different numbers of $n_{\text{treat}}$.
Each optimizer is applied on each scenario with 20 stochastic repetitions.

Figure~\ref{fig:plot_boxplot_valid_y} shows the estimated power values $y_{\text{valid}}$ for the best solutions in each scenario.
\begin{figure}[htb]
\centering
\includegraphics[width=\linewidth]{generated/figures/plot_boxplot_valid_y.pdf}
\caption{%
  Box plots of performance (power) values on independent replicates of the best solutions for the four optimizers, the different effect sets, and different numbers of $n_{\text{treat}}$
  }
\label{fig:plot_boxplot_valid_y}
\end{figure}
The plots show that \emph{BO} generally gives results that are comparable to those obtained with the (exhaustive) \emph{Grid} method across all scenarios. 
Only in some cases \emph{Grid} has a slight advantage (\emph{effect: linear}, $n_{\text{treat}} = 2000$; \emph{effect: sigmoid}, $n_{\text{treat}} = 1000$; \emph{effect: sigmoid}, $n_{\text{treat}} = 2000$).
Below we will see that these scenarios have in common that the optimal configuration is close to the border of the search space. 
In other cases \emph{BO} yields slightly better results (\emph{effect: paper}, $n_{\text{treat}} = 2000$; \emph{effect: paper2}, $n_{\text{treat}} = 500$; \emph{effect: sigmoid}, $n_{\text{treat}} = 500$).
Across all scenarios, \emph{BO} and \emph{Grid} perform similar, even though \emph{BO} due to its evaluation budget is allowed only 116 evaluations in contrast to the 1350 evaluations for \emph{Grid}, with a resolution of 25 points per dimension. 
The difference in the number of evaluations is clearly reflected in the average runtime of the optimization methods given in Table~\ref{tab:table_time}.
\emph{BO} is approximately 20 times faster.
\input{generated/tables/table_time.tex}

For \emph{BO Grid}, only solutions on the grid that was evaluated by \emph{Grid} are considered in the evaluation. 
Comparing the results of \emph{BO Grid} to those of \emph{BO} we mainly see similar performances for both.
This indicates, that the resolution of \emph{Grid} is fine enough to achieve performance similar to \emph{BO}.

>>>>

\title[Bayesian optimization of Adaptive Seamless Designs]{Improving Adaptive Seamless Designs through Bayesian optimization} %(please use sentence case, e.g. Score tests for exploring complex models)
%% Information for the first author.
\author[Jakob Richter {\it{et al.}}]{Jakob Richter\footnote{Corresponding author: {\sf{e-mail: richter@statistik.tu-dortmund.de}}}\inst{,1}} 
\address[\inst{1}]{Fakultät Statistik, Technische Universität Dortmund, 44221 Dortmund}
%%%%    Information for the second author
\author[dd]{Tim Friede\inst{2}}
\address[\inst{2}]{Institut für Medizinische Statistik, Universitätsmedizin Göttingen, and Deutsches Zentrum für Herz-Kreislauf-Forschung (DZHK), Standort Göttingen, 37073 Göttingen}
%%%%    Information for the third author
\author[]{Jörg Rahnenführer\inst{1}} %please provide full author names. Middle names should be indicated by initials only, i.e. Henry J. James
%%%%    \dedicatory{This is a dedicatory.}
\Receiveddate{zzz} \Reviseddate{zzz} \Accepteddate{zzz}

@@

% max 250 words
\begin{abstract}
In this paper we propose to use Bayesian optimization (BO) to improve the efficiency of the selection process of designs in clinical trials.
BO is a method to optimize expensive black-box functions that uses a regression as a surrogate to guide the search.
In clinical trials, planning test procedures and sample sizes is a crucial task.
A common goal is to maximize the test power, given a set of treatment and effect sizes under the null hypothesis.
From a wide range of possible alternatives we want to select the best plan in a short period of time, to enable quick decisions.
The standard approach to simulate the power for each single setup can become too time-consuming, when the number of possible setups becomes very large.
Then, either large computational resources are required or an exhaustive exploration of all possible setups takes a very long time.
In this work we propose to use BO to quickly find a clinical trial setup with high power from a large number of setups options.
We demonstrate the effectiveness of our approach by optimizing the power of adaptive seamless designs for different sets of treatment effect sizes.
We compare the BO approach with an exhaustive evaluation of all candidate designs.
The optimization approach finds competitive setups in a fraction of the time.
\end{abstract}

@@

% FIXME: Do we really take in this Assurance?
% Assurance: \cite{stallard_optimal_2009} in diskussion abhandeln

@@

We start by providing some more details on the study design considered (Sections \ref{ssec:two_stage_design_with_treatment_selection} and \ref{ssec:selection_rules}) and then introduce briefly a framework for optimizing such designs, namely clinical scenario evaluation (CSE) in Section \ref{ssec:clinical_scenario_evaluation_for_asd}. %In Section \ref{ssec:simulating_asd} we make some brief comments on the simulation model.

\subsection{Two-stage design with treatment selection}
\label{ssec:two_stage_design_with_treatment_selection}
Here we consider a particular type of adaptive seamless designs, namely two-stage designs with treatment (or more specifically dose) selection. The trial starts in stage 1 with $K$ experimental treatments or doses of one experimental treatment which are compared to a common control. Interest is in assessing the treatment effects $\beta_k$ with $k=1,\dots, K$ comparing experimental treatment $k$ vs. control and testing the null hypotheses $H_{0,k}: \beta_k=0$ against the alternative  $H_{1,k}: \beta_k>0$ (larger $\beta_k$ being better). For the purpose of treatments selection, test statistics $Z_{1,k}$ based on data from the first design stage are considered with larger $Z_{1,k}$ indicating more beneficial treatments. These can use data on the primary outcome used for testing or any other outcome that is available at the time of the interim analysis. In practical applications, early outcomes are often used to inform interim decisions, since the primary endpoint might take a longer time to observe. In the interim analysis, treatments are selected for continuation into the second stage; the set of selected treatments is denoted by $\mathcal{S}$. The selection rules used here are described in Section \ref{ssec:selection_rules}. In the final analysis, the closed test principle is applied to account for the multiple hypotheses. The intersection hypotheses are tested by combining stagewise $p$-values for the respective intersection hypotheses using a prespecified combination function. Here we use the inverse normal combination function~\citep{lehmacher_adaptive_1999} which is given by
\begin{equation}
   w_1 \Phi^{-1}(1-p_{\mathcal{K},1}) + w_2 \Phi^{-1}(1-p_{\mathcal{K},2}) \,
\end{equation}
where $\Phi^{-1}(\cdot)$ denotes the quantile function of the standard normal distribution, $w_1$ and $w_2$ weights with $w_1^2+w_2^2 = 1$, and $p_{\mathcal{K},1}$ and $p_{\mathcal{K},2}$ stagewise $p$-values testing intersection hypothesis $H_{\mathcal{K}}$. The p-values for stages $1$ and $2$ are calculated on those patient recruited in stages $1$ and $2$, respectively. Note that one might not be able to calculate $p_{\mathcal{K},1}$ at the time of the interim analysis since some patients might still be under follow-up for the final endpoint.

\subsection{Selection rules}
\label{ssec:selection_rules}
In practice, the totality of the data would be considered by a data monitoring committee (DMC) to make a recommendation regarding the treatment selection. However, it is advisable for the sponsor to define a selection rule, although in comparison fairly simplistic, to provide some guidance for the DMC regarding the sponsor's preferences. Furthermore, formal selection rules are indispensable to conduct simulation studies, e.g. to explore sample size and power. In the following we give some examples of selection rules which have previously been consider and which will feature below.

The treatment selection might consider statistics $Z_{1,k}$ for the final (primary) outcome or statistics $Z_{1,k}^{(E)}$ calculated based on some early outcome. We present the selection rules using $Z_{1,k}^{(E)}$, but of course these would be replaced by $Z_{1,k}$ if the final outcome was used for treatment selection. Furthermore, larger statistics are considered better.

\paragraph{$\kappa$-best Rule} Let $Z_{1,(1)}^{(E)}, \dots, Z_{1,(K)}^{(E)}$ denote the ordered statistics with $Z_{1,(1)}^{(E)}$ being the largest and therefore best test statistic. The $\kappa$-best rule then selects the treatments associated with the $\kappa$ largest $Z_{1,k}^{(E)}$, i.e.$Z_{1,(1)}^{(E)}, \dots, Z_{1,(\kappa)}^{(E)}$.

\paragraph{Epsilon Rule} To our knowledge this rule was first described by \citet{kelly_adaptive_2005} and has since then featured in a number of simulation studies including \citet{friede_comparison_2008, friede_adaptive_2020}. All treatments $k$ with $Z_{1,k}^{(E)} \ge \max_i Z_{1,i}^{(E)} - \epsilon$ with $\epsilon \ge 0$ are selected. With $\epsilon=0$ this rule is equivalent to the $\kappa$-best rule with $\kappa=1$.

\paragraph{Threshold Rule} With this rule all treatments $k$ are selected with $Z_{1,k}^{(E)} \ge \tau$ where $\tau$ is a given threshold.

We note that with the $\kappa$-best rule a fixed number of treatments is carried forward into the next stage whereas these are variable with the epsilon rule and the threshold rule.

@@

The clinical scenario evaluation framework was first proposed by \citet{benda_aspects_2010} and subsequently developed further by \citet{friede_refinement_2010}. 
Figure \ref{fig:friede_refinement_2010_fig2} provides a summary of this framework.
\begin{figure}[ht]
\centering
\begin{tikzpicture}[auto,node distance=1cm,align=center,text width = 3cm, inner sep = 0.1cm, font=\sffamily \scriptsize, outer sep= 0cm]
  \node[draw, ellipse, minimum height=1cm, text width=2.5cm] (ClinicalScenarios) at (0,0) {
    \textbf{Clinical scenarios}\\
    Simulation Studies};
  \node[name=block, rectangle split, rectangle split parts=2, draw, above = of ClinicalScenarios] (UnknownPoints)
  { \begin{itemize}[leftmargin=.3cm, topsep=0cm,itemsep=0cm,partopsep=0cm, parsep=0cm]
      \item Treatment effect on early interim outcome
      \item Treatment effect on late disability outcome
    \end{itemize}
    \nodepart{second} Unknown
  };
  \node[name=block, rectangle split, rectangle split parts=2, draw, anchor = north east] (HistoricalDataPoints) at (UnknownPoints.north west)
  { \begin{itemize}[leftmargin=.3cm, topsep=0cm,itemsep=0cm,partopsep=0cm, parsep=0cm]
      \item Correlation between early interim and late disability outcomes
    \end{itemize}
    \nodepart{second} Historical data
  };
  %\node[draw, inner xsep=1em, inner ysep=1em, above left = of ClinicalScenarios, fit = (HistoricalDataPoints) (UnknownPoints)] (boxDiseaseSpecific) {};
  \node[draw, text width = 6.2cm, anchor = south west] (DiseaseSpecific) at (HistoricalDataPoints.north west) {\textbf{Disease-specific features}};
  %\node[draw, above right = of ClinicalScenarios] (DesignOptions) {Design options};
  \node[name=block, rectangle split, rectangle split parts=2, draw, anchor = north west] (ConstrainedPoints) at ([xshift = 4cm]HistoricalDataPoints.north east)
  { \begin{itemize}[leftmargin=.3cm, topsep=0cm,itemsep=0cm,partopsep=0cm, parsep=0cm]
      \item Number of test treatments
      \item Total subjects in trial
      \item Recruitment rate
    \end{itemize}
    \nodepart{second} Constrained by health care environment/infrastructure
  };
  \node[name=block, rectangle split, rectangle split parts=2, draw, anchor = north west] (CombinationPoints) at (ConstrainedPoints.north east)
  { Adaptive seamless design
    \begin{itemize}[leftmargin=.3cm, topsep=0cm,itemsep=0cm,partopsep=0cm, parsep=0cm]
      \item Time point of interim analysis
      \item Treatment selection rules at interim 
      \item Conventional phase 2/3
      \item Split of resources between phases
      \item End oh phase futility criteria
    \end{itemize}
    \nodepart{second} Combination of options
  };
  \node[draw,text width = 6.2cm, anchor = south west] (DesignOptions) at (ConstrainedPoints.north west) {\textbf{Design Options}};
  \node[draw, below = 0.5cm of ClinicalScenarios] (DesignPerformance) {
    \textbf{Design performance} \\ e.g.\ statistical power};
  \node[draw, right = of DesignPerformance, text width = 4cm, node distance = 2cm] (Evaluation) {
    \textbf{Clinical Scenario Evaluation}\\
    Design performance measures evaluated across a wide range of clinical scenarios};
  \draw[->] (HistoricalDataPoints) -- (ClinicalScenarios);
  \draw[->] (UnknownPoints) -- (ClinicalScenarios);
  \draw[->] (CombinationPoints.223) -- (ClinicalScenarios);
  \draw[->] (ConstrainedPoints) -- (ClinicalScenarios);
  \draw[->] (ClinicalScenarios) -- (DesignPerformance);
  \draw[->] (DesignPerformance) -- (Evaluation);
\end{tikzpicture}
%Friede et al (2010) DIJ, Figure 2 https://link.springer.com/content/pdf/10.1177/009286151004400507.pdf? https://link.springer.com/content/pdf/10.1177/009286151004400607.pdf
\caption{Application of the refined clinical scenario evaluation framework to adaptive seamless designs in progressive multiple sclerosis (Figure adapted from \citet{friede_refinement_2010}).}
\label{fig:friede_refinement_2010_fig2}
\end{figure}
A clinical scenario evaluation is defined by the following three elements: (i) disease-specific features; (ii) design options; and (iii) design performance measures. The design specific features describe distributional assumptions on the early and final outcomes including variances, correlations and treatment effect sizes. Here, those that can be estimated from data of previous clinical trials and those that are simply unknown. Considering adaptive seamless design, the design options include the time point of the interim analysis and the treatment selection rule, which can in principle both be chosen without any restrictions. In contrast, some other design options such as the number of treatments and the total sample size are constrained by the environment and infrastructure available. The performance measures (sometimes also referred to as metrics) are in the context with adaptive seamless designs e.g. power or sample size distributions.
%\paragraph{Simulating ASD}
%\label{ssec:simulating_asd}
The evaluation of some performance measures requires Monte Carlo simulations, since no closed form solutions are available. 
The R-package \texttt{asd} by~\citet{parsons_software_2011} can be used for such simulations.
As it generates test statistics it is generally more efficient in comparison to simulating individual participant data.
The simulation model was described in more detail by \citet{friede_adaptive_2020}. 
It features selection at interim based on early as well as final outcomes.

% Some corrections to the figure needed: e.g. design options (right panel)

@@

@@

The strategies for treatment selection for the second stage in the two-phase trial design and their corresponding parameters that are used in the simulation study are given in Table~\ref{tab:selection_strategies} and Table~\ref{tab:search_space}, respectively.
The strategies follow the concept of adaptive seamless designs discussed in Section~\ref{sec:adaptive_seamless_designs}. 
The strategies 1, 2, 3, and \emph{all} select the $k$ best strategies according to their respective test statistics in the first stage (with \emph{all} representing all $k=4$ treatments besides the control). 
The strategies \emph{eps} and \emph{thresh} represent strategies with a flexible number of treatments taken to the second stage. 
In Table~\ref{tab:search_space} the ranges for these two parameters and for the proportion of patients (ratio $r$) allocated to the second stage are given.
\begin{table}[h]
  \caption{Overview of selection strategies to determine which trial arms are kept for second stage}
  \label{tab:selection_strategies}
  \centering
  \begin{tabular}{ll}
  \hline
  Notation       & Procedure of selection strategy  \\
  \hline
  \emph{$\kappa$-best}    & Select $\kappa = 1, 2, 3$ or 4 (all) arms, respectively, according to maximal test statistic \\
  \emph{eps}     & Select all arms with test statistic less than $\epsilon$ (epsilon) below maximal test statistic \\
  \emph{thresh}  & Select all arms with test statistic larger than the value $\tau$ (threshold) \\
  \hline
  \end{tabular}
\end{table}
\begin{table}[h]
  \caption{Parameters for the selection strategies representing the search space of trial designs}
  \label{tab:search_space}
  \centering
  \begin{tabular}{ll}
  \hline
  Parameter                        & Range \\
  \hline
  Selection strategy               & \emph{\{1-best, 2-best, 3-best, all, eps, thresh\}} \\
  $r$ (ratio)                      & $(0,1)$ \\
  $\epsilon$ (only for \emph{eps}) & $[0,4]$ \\
  $\tau$ (only for \emph{thresh})  & $[0,10]$ \\
  \hline
  \end{tabular}
\end{table}

@@

\paragraph{Grid and Grid small}
We conduct an \emph{exhaustive grid search} with a resolution of $k$ points per real-valued dimension and one point for each categorical dimension.
Referring to our application, for each selection strategy we evaluate a numerical grid with a resolution of $k$.
All 6 selection strategies have the numeric parameter $r$.
The strategies \emph{epsilon} and \emph{thresh} both have an additional second parameter, $\epsilon$ or $\tau$, respectively.
Using $k=25$, we call this approach \textbf{Grid},
with in a total $4 \cdot 25 + 2 \cdot 25 \cdot 25 = 1350$ design configurations that are evaluated for one scenario.
Using a subset of this grid with $k=7$ points per dimension, resulting in $4 \cdot 7 + 2 \cdot 7 \cdot 7 = 126$ design configurations, we call the method \textbf{Grid Small}.
Both grids are evaluated with 20 stochastic replications to estimate the variance of the estimated power.

@@

\paragraph{BO (Model-based optimization) and BO Grid}
For applying BO, we first define how the surrogate model that predicts the outcome of $f$ for unknown values of $\theta$ (parameter configurations) is generated.
As regression model for the surrogate we choose Kriging (also called Gaussian process regression) and use the implementation in the R-package \texttt{DiceKriging}~\citep{roustant_dicekriging_2012}, because it is known to work well for fairly low dimensional search spaces.
We configure Kriging to apply the Mattern$\frac{5}{2}$ kernel with an estimated \emph{nugget effect} to account for the noisy response of $f$ and without scaling the input variables to $[0,1]$.

@@

The whole BO procedure is repeated $20$ times to estimate the variance of the optimal solution found.

@@

The algorithms used in this study are implemented in R.
For the Bayesian optimization, the R-package \texttt{mlrMBO}~\citep{bischl_mlrmbo_2017} is used.
For the simulation of clinical trials, the function \texttt{treatsel.sim} of the R-package \texttt{asd}~\citep{parsons_software_2011} is used, with parameter values given in Table~\ref{tab:par_implement}.
The vector of treatment numbers for determining power \texttt{ptest} counts rejections of 3 or 4 hypotheses ???.
Note that the parameter \texttt{nsim} denotes the internal simulation iterations of $f$. 
The ten additional stochastic repetitions we conduct for the \emph{grid search} and \emph{BO} are additional replicates.

@@

\begin{table}[h]
  \caption{Parameter choices for the R function \texttt{treatsel.sim}}
  \label{tab:par_implement}
  \centering
  \begin{tabular}{lll}
  \hline
  Meaning of parameter & Parameter & Value \\
  \hline
  Number of simulation iterations & \texttt{nsim} & $1000$ \\
  Correlation between early and final outcomes & \texttt{corr} & $0.4$ \\
  Test level & \texttt{level} & $0.025$ \\
  Vector of treatment numbers for determining power & \texttt{ptest} & $(3,4)$ \\
  \hline
  \end{tabular}
\end{table}

\subsection{Main results for comparison of algorithms}

In this chapter we present an overview of the results for the comparison of the four optimization strategies: exhaustive \emph{Grid} search, grid search with a reduced resolution (\emph{Grid Small}), \emph{BO} and \emph{BO Grid}.
We focus our comparison on independent, replicated evaluations of the configurations proposed by each optimizer.
The evaluations provide unbiased estimates of the performance ($y_{\text{valid}}$) of all optimizers.

For this study we consider 12 scenarios.
The scenarios are generated by combining the four different effect sets (see Table~\ref{tab:table_effect_names}) with the three different numbers for $n_{\text{treat}}$.
Each optimizer is applied on each scenario with 20 stochastic repetitions.

Figure~\ref{fig:plot_boxplot_valid_y} shows the estimated power values $y_{\text{valid}}$ for the best solutions in each scenario.
The plots show that \emph{BO} generally gives results that are comparable to those obtained with the (exhaustive) \emph{Grid} method across all scenarios. 
Only in some cases \emph{Grid} has a slight advantage (\emph{effect: linear}, $n_{\text{treat}} = 2000$; \emph{effect: sigmoid}, $n_{\text{treat}} = 1000$; \emph{effect: sigmoid}, $n_{\text{treat}} = 2000$).
Below we will see that these scenarios have in common that the optimal configuration is close to the border of the search space. 
In other cases \emph{BO} yields slightly better results (\emph{effect: paper}, $n_{\text{treat}} = 2000$; \emph{effect: paper2}, $n_{\text{treat}} = 500$; \emph{effect: sigmoid}, $n_{\text{treat}} = 500$).
Across all scenarios, \emph{BO} and \emph{Grid} perform similarly, although due to its evaluation budget, \emph{BO} is allowed only 116 evaluations as opposed to the 1350 evaluations for \emph{Grid}, when using a resolution of 25 points per dimension.
The difference in the number of evaluations is clearly reflected in the average runtime of the optimization methods given in Table~\ref{tab:table_time}.
\emph{BO} is approximately 20 times faster.
\input{generated/tables/table_time.tex}

\begin{figure}[tbh]
\centering
\includegraphics[width=\linewidth]{generated/figures/plot_boxplot_valid_y.pdf}
\caption{%
  Box plots of performance (power) values on independent replicates of the best solutions for the four optimizers, the different effect sets, and different numbers of $n_{\text{treat}}$ (500, 1000, 2000)
  }
\label{fig:plot_boxplot_valid_y}
\end{figure}

For \emph{BO Grid}, only solutions on the grid that was evaluated by \emph{Grid} are considered in the evaluation. 
Comparing the results of \emph{BO Grid} to those of \emph{BO} we mainly see similar performances for both.
This indicates that the resolution of \emph{Grid} is fine enough to achieve similar performance to \emph{BO}.
